% conclusion.tex — HapGraph
% Target: 200+ words

We have presented \hapgraph{}, an open-source Python package implementing
a two-stage sequential pipeline for admixture graph inference and IBD-based
timing estimation from phased SNP data.
Stage~I uses F-statistics to infer the graph topology and estimate admixture
proportions; Stage~II fixes that topology and estimates the admixture timing
$T$ via Bayesian NUTS-MCMC over an IBD segment-length likelihood.
This design addresses a gap in population genetic practice: the lack of a
unified tool that connects the \emph{who} (topology and proportions) with
the \emph{when} (timing), while remaining computationally tractable at
$N \gtrsim 10$ populations.
We emphasise that \hapgraph{} is a sequential pipeline, not a fully joint
Bayesian inference: the topology is treated as fixed in Stage~II and is
not a random variable in the posterior.

Simulation experiments across three scenarios demonstrate that \hapgraph{}
achieves zero false positive rate on pure trees; greedy topology recall
reaches 85\% for $K=1$ (S2) and 97.5\% for $K=2$ (S3), with precision 85\%
and 100\% respectively.
Under oracle topology, the $T$ posterior produces correct coverage
(95--97.5\%) but with mean CI widths of $\sim$43 generations, indicating
that timing is not precisely identified in the 0.5-Morgan simulation
benchmark; the model is correctly specified but underpowered to resolve
$T$ in this setting.
The key limitation identified in benchmarking — severely under-calibrated
uncertainty for admixture proportions (10--20\% empirical 95\% CI coverage)
— is attributable to the delta-method approximation breaking down under high
genetic drift ($N_e = 1{,}000$) and near $\alpha \approx 0.5$; improved
uncertainty quantification for $\alpha$ is a priority for future development.

Application to the 1000 Genomes Project 2022 dataset yielded eight admixture
events; seven of the eight targets (ASW, ACB, PUR, CLM, MXL, BEB, PJL)
are populations whose admixed ancestry is well established in the literature,
providing external validation of the topology inference.
The eighth inferred event (LWK) carries a significant $F_3$ Z-score but
lacks a clear single-pulse biological interpretation and is flagged for
further validation (see Results).
These results demonstrate that \hapgraph{}'s topology inference scales to
26 populations and identifies known admixture events without manual curation.

Future directions include: extending simulation validation to $K \geq 3$;
applying IBD-based timing estimation to real data following hap-ibd
pre-processing; improving alpha credible interval calibration; and direct
quantitative comparison to \textsc{GLOBETROTTER} and \alder{} timing estimates.
\hapgraph{} is designed as a modular Python package, allowing components
(F-statistics, IBD model, MCMC sampler) to be extended or replaced
independently, facilitating community development and adaptation to new
use cases.
